Drought indices of water demand and supply, soil moisture, vegetation, and its impact on LULCC in continental Chile

### 2024-01-18 article/paper_drought_Chile2023.tex
update mañana 17 ene

\begin{abstract}
Human-induced greenhouse gas emissions have increased the frequency
and/or intensity of weather and climate extremes. Central Chile has been
affected by a persistent drought which is impacting the hydrological
system and vegetation development. The region has been the focus of
research studies due to the diminishing water supply, this persistent
period of water scarcity has been defined as a ``mega drought''.
Nevertheless, our results evidence that the water deficit has expanded
beyond. Our goal is to analyze the impact of drought, measured by
drought indices of water supply/demand and vegetation status, in the
LULCC (land use land cover change) over continental Chile. For the
analysis, continental Chile was divided into five zones according to a
latitudinal gradient: ``Norte Grande'', ``Norte Chico'', ``Zona
Central'', ``Zona Sur'', and ``Zona Austral''. We used monthly climatic
re-analysis variables for precipitation, temperature and soil moisture
for 1981-2023 from ERA5-Land; and MODIS (Moderate-Resolution Imaging
Spectroradiometer) product MCD12Q1 for land cover for 2001-2021, and the
NDVI vegetation index from product MOD13A2 collection 6.1 for 2000-2023,
both from collection 6.1. We estimated atmospheric evaporative demand
(AED) by combining the Hargreaves-Samani equation with the ERA5-Land
temperature. We derived the drought indices SPI (Standardized
Precipitation Index), SPEI (Standardized Precipitation
Evapotranspiration Index), EDDI (Evaporative Demand Drought Index), SSI
(standardized anomaly of cumulative soil moisture), and the zcNDVI
(standardized anomaly of cumulative NDVI). These indices were calculated
for time scales of 1, 3, 6, 12, 24, and 36 months, except for zcNDVI (1,
3, and 6 months). We analyzed the temporal correlation of SPI, SPEI,
EDDI, and SSI with zcNDVI to have insights into the impact of water
supply and demand on vegetation. Our results showed that LULCC had an
increasing trend of 412 {[}km2yr−1{]} of forest expansion in the ``Zona
Sur'', together with a decreasing trend of 24 {[}km2yr−1{]} of cropland
contraction in the ``Zona Central'' meanwhile the ``Zona Sur'' showed an
increase of 31 {[}km2yr−1{]}, and a contraction of 80 {[}km2yr−1{]} of
bare soil in the ``Zona Austral''. The EDDI was the less correlated
index for the five macro zones and the five types of land cover, showing
that the temperature in Chile has little impact on vegetation. Higher
r-squared values, between 0.5 and 0.8, were obtained at ``Norte Chico''
and ``Zona Central'' for the land cover types of savanna, shrubland,
grassland, and croplands for the indices SPEI and SSI at time scales of
12 and 24 months. The forest type reaches a r-squared of
\textasciitilde0.5 for SSI of 12 months. The results indicate that the
``Norte Chico'' and ``Zona Central'' are the most sensitive regions to
water supply deficits longer than a year, potentially explained by a low
capacity of water storage in those zones that should be further
investigated.
\end{abstract}

@@

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, interior hidden, enhanced, frame hidden, breakable, sharp corners, boxrule=0pt]}{\end{tcolorbox}}\fi

@@

In this research, we present the raster dataset DDS4Chl, which provides
climate variables and drought indices of water demand and supply and
vegetation productivity for continental Chile since 1981 at a monthly
frequency. Those were gathered from the earth observation products
ERA5-Land and MODIS. Then, we used DDS4Chl to analyze the impact of
drought on different types of land cover classes in continental Chile.
The specific objectives of the study are: i) to analyze the trend on
multi-scalar drought indices for water demand and supply, soil moisture,
and vegetation productivity for 1981--2023/2001--2023; ii) to evaluate
LULC change for 2001--2021 and its relation to drought indices; iii) to
analyze the relationship of a proxy of vegetation productivity (zcNDVI)
with drought indices of water demand and supply and soil moisture; and
iv) to assess if the observed changes in the drought indices are linked
to TWS.

\hypertarget{study-area}{%
\section{Study area}\label{study-area}}

@@

For the downloading, processing, and analysis of the spatio-temporal
data, we used the open source software for statistical computing and
graphics, \texttt{R} \citep{R2023}. For downloading ERA5-Land, we used
the \texttt{\{ecmwfr\}} package \citep{Hufkens2019}. For processing
raster data, we used \texttt{\{terra\}} \citep{Hijmans2023} and
\texttt{\{stars\}} \citep{Pebesma2023}. For managing vectorial data, we
used \texttt{\{sf\}} \citep{Pebesma2018}. For the calculation of AED, we
used \texttt{\{SPEI\}} \citep{Bergueria2023}.

@@

For water supply and demand variables, we used ERA5-Land
\citep{MunozSabater2021}, a reanalys dataset that provides the evolution
of land variables since 1950. It has a spatial resolution of 0.1°,
hourly frequency, and global coverage. We selected the variables for
total precipitation, 2 meter temperature maximum and minimum, and
volumetric soil water layers between 0 and 100cm of depth (layer 1 to
layer 3). The data was downloaded using the Copernicus Climate Data
Store (CDS) Application Program Interface (API) implemented in
\texttt{\{ecmfwr\}} \citep{Hufkens2019}.

@@

\begin{table}[!ht]
\caption{Description of the earth observation data used }
\label{tab-desEOD}
\small
\centering
\begin{tabular}{p{0.13\textwidth}cp{0.3\textwidth}p{0.095\textwidth}ccc}
\hline
\multirow{1}{*}{\centering Product} & Sub-product & Variable & Spatial Resolution  & Period & Units & Short Name \\ 
\hline
\multirow{4}{*}{ERA5-Land} & ~ & Precipitation & \multirow{4}{*}{~0.1°} & \multirow{4}{*}{1981-2023} & mm & P \\ 
         &  & Maximum temperature & ~ & & $°C$ & $T_{max}$ \\ 
         &  & Minimum temperature & ~ & & $°C$ & $T_{min}$ \\ 
         &  & Volumetric Soil Water Content at 1m & ~ & & $m3/m3$ & SM \\ 
ERA5-Land* & & Atmospheric Evaporative Demand & 0.1° & 1981-2023 & mm & AED \\
        \multirow{2}{*}{MODIS} & MOD13A3.061 & Normalized Difference Vegetation Index & \multirow{2}{*}{~1 km} & 2000-2023 & ~ & NDVI \\ 
         & MCD12Q1.061 & Landcover IGBP scheme & & 2001-2022 & ~ & Landcover \\ 
\hline
\end{tabular}
{\raggedright *Derived from ERA5-Land with Eq. \ref{eq-AED}. \par}
\end{table}

@@

\hypertarget{validation-of-era5-land-variables}{%
\subsection{Validation of ERA5-Land
variables}\label{validation-of-era5-land-variables}}

@@

Figure~\ref{fig-zcNDVI_var} and \ref{tab-landcoverTrend} show that the
macrozones with major LULCC for 2001-2022 were ``Centro'', ``Sur'', and
``Austral'' with 36\%, 31\%, and 34\%, respectively. The ``Norte Chico''
shows an increase in barrend land of 111 \(km^2 year^{-1}\) and a
reduction in the class savanna -70 \(km^2 year^{-1}\). In the ``Centro''
and ``Sur'' there are changes in the chilean matorral, with an important
reduction in savanna (-136 to -318 \(km^2 yr^{-1}\)), and an increase in
shrubland and grassland. Showing a change for more dense vegetation
types. It appears to be a change from the surface occupied by cropland
from the ``Centro'' to the ``Sur.'' Also, there is a high increase in
forest (397 \(Km^2 yr^{-1}\)) in the ``Sur,'' replacing the savanna
lost.

@@

Regarding vegetation productivity aggregated through the macrozones in
the five landcover macroclasses. In ``Norte Grande'' there is a increase
trend of 0.02 (z-index) per decade, related with types of grassland and
shrubland. There is a negative trend in ``Norte Chico'' with -0.04 and
``Centro'' with -0.02 per decade. In the ``Norte Chico,'' savanna
(-0.05) has the lowest trend, and the rest of the types are around
-0.04. In ``Centro,'' shrubland reaches -0.06, followed by grassland
with -0.05, and croplands and savanna have \textasciitilde-0.01 per
decade. This could be associated either with a reduction in vegetation
surface, a decrease in biomass, or browning \citep{Miranda2023}.
Vegetation reached its lowest values since the year 2019, reaching an
extreme condition in early 2020 and 2022 in the ``Centro'' (Mega
Drought). The ``Sur'' and ``Austral'' show a positive trend of around
0.016 per decade (Figure~\ref{fig-zcNDVI_var}). The area most affected
by a negative trend in vegetation seems to be associated with the
Chilean matorral \citep{Fuentes2021}, despite the croplands being as
severely impacted by drought as native vegetation does in ``Norte
Chico.''.

@@

The correlation between zcNDVI-6 and drought indices shows that time
scales higher than six months are predominant over continental Chile in
explaining vegetation variability (Figure~\ref{fig-corTimeScale}).
Pearson correlations between 0.6 and 0.9 are concentrated in the ``norte
chico'' and ``zona central'' for SPI and SPEI. Soil moisture reached a
correlation of 0.6--0.9 that extended north and south with respect to
SPI.

\hypertarget{impact-for-water-supply-and-demand-and-soil-moisture-in-vegetation-productivity}{%
\subsection{Impact for water supply and demand, and soil moisture in
vegetation
productivity}\label{impact-for-water-supply-and-demand-and-soil-moisture-in-vegetation-productivity}}

\begin{table}[!ht]
\label{tab-corLandcover}
\caption{}
\includegraphics[]{../output/figs/tabla_r_cor_macro_indice.png}
\end{table}

@@

\hypertarget{validation-of-era5-land-variables-1}{%
\subsection{Validation of ERA5-Land
variables}\label{validation-of-era5-land-variables-1}}

@@

>>>>

\begin{abstract}
Human-induced greenhouse gas emissions have increased the frequency
and/or intensity of weather and climate extremes. Central Chile has been
affected by a persistent drought which is impacting the hydrological
system and vegetation development. The region has been the focus of
research studies due to the diminishing water supply, this persistent
period of water scarcity has been defined as a ``mega drought''.
Nevertheless, our results evidence that the water deficit has expanded
beyond. Our goal is to analyze the impact of drought, measured by
drought indices of water supply/demand and vegetation status, in the
LULCC (land use land cover change) over continental Chile. For the
analysis, continental Chile was divided into five zones according to a
latitudinal gradient: ``Norte Grande'', ``Norte Chico'', ``Zona
Central'', ``Zona Sur'', and ``Zona Austral''. We used monthly climatic
re-analysis variables for precipitation, temperature and soil moisture
for 1981-2023 from ERA5-Land (ERA5L); and MODIS (Moderate-Resolution
Imaging Spectroradiometer) product MCD12Q1 for land cover for 2001-2021,
and the NDVI vegetation index from product MOD13A2 collection 6.1 for
2000-2023, both from collection 6.1. We estimated atmospheric
evaporative demand (AED) by combining the Hargreaves-Samani equation
with the ERA5L temperature. We derived the drought indices SPI
(Standardized Precipitation Index), SPEI (Standardized Precipitation
Evapotranspiration Index), EDDI (Evaporative Demand Drought Index), SSI
(standardized anomaly of cumulative soil moisture), and the zcNDVI
(standardized anomaly of cumulative NDVI). These indices were calculated
for time scales of 1, 3, 6, 12, 24, and 36 months, except for zcNDVI (1,
3, and 6 months). We analyzed the temporal correlation of SPI, SPEI,
EDDI, and SSI with zcNDVI to have insights into the impact of water
supply and demand on vegetation. Our results showed that LULCC had an
increasing trend of 412 {[}km2yr−1{]} of forest expansion in the ``Zona
Sur'', together with a decreasing trend of 24 {[}km2yr−1{]} of cropland
contraction in the ``Zona Central'' meanwhile the ``Zona Sur'' showed an
increase of 31 {[}km2yr−1{]}, and a contraction of 80 {[}km2yr−1{]} of
bare soil in the ``Zona Austral''. The EDDI was the less correlated
index for the five macro zones and the five types of land cover, showing
that the temperature in Chile has little impact on vegetation. Higher
r-squared values, between 0.5 and 0.8, were obtained at ``Norte Chico''
and ``Zona Central'' for the land cover types of savanna, shrubland,
grassland, and croplands for the indices SPEI and SSI at time scales of
12 and 24 months. The forest type reaches a r-squared of
\textasciitilde0.5 for SSI of 12 months. The results indicate that the
``Norte Chico'' and ``Zona Central'' are the most sensitive regions to
water supply deficits longer than a year, potentially explained by a low
capacity of water storage in those zones that should be further
investigated.
\end{abstract}

@@

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, frame hidden, sharp corners, breakable, borderline west={3pt}{0pt}{shadecolor}, interior hidden, boxrule=0pt]}{\end{tcolorbox}}\fi

@@

In this research, we present the raster dataset DDS4Chl, which provides
climate variables and drought indices of water demand and supply and
vegetation productivity for continental Chile since 1981 at a monthly
frequency. Those were gathered from the earth observation products ERA5L
and MODIS. Then, we used DDS4Chl to analyze the impact of drought on
different types of land cover classes in continental Chile. The specific
objectives of the study are: i) to analyze the trend on multi-scalar
drought indices for water demand and supply, soil moisture, and
vegetation productivity for 1981--2023/2001--2023; ii) to evaluate LULC
change for 2001--2021 and its relation to drought indices; iii) to
analyze the relationship of a proxy of vegetation productivity (zcNDVI)
with drought indices of water demand and supply and soil moisture; and
iv) to assess if the observed changes in the drought indices are linked
to TWS.

\hypertarget{study-area-agregar-superficie-de-landcover-en-cada-macrozona}{%
\section{Study area (AGREGAR SUPERFICIE DE LANDCOVER EN CADA
MACROZONA)}\label{study-area-agregar-superficie-de-landcover-en-cada-macrozona}}

@@

For the downloading, processing, and analysis of the spatio-temporal
data, we used the open source software for statistical computing and
graphics, \texttt{R} \citep{R2023}. For downloading ERA5L, we used the
\texttt{\{ecmwfr\}} package \citep{Hufkens2019}. For processing raster
data, we used \texttt{\{terra\}} \citep{Hijmans2023} and
\texttt{\{stars\}} \citep{Pebesma2023}. For managing vectorial data, we
used \texttt{\{sf\}} \citep{Pebesma2018}. For the calculation of AED, we
used \texttt{\{SPEI\}} \citep{Bergueria2023}.

@@

For water supply and demand variables, we used ERA5L
\citep{MunozSabater2021}, a reanalys dataset that provides the evolution
of land variables since 1950. It has a spatial resolution of 0.1°,
hourly frequency, and global coverage. We selected the variables for
total precipitation, 2 meter temperature maximum and minimum, and
volumetric soil water layers between 0 and 100cm of depth (layer 1 to
layer 3). The data was downloaded using the Copernicus Climate Data
Store (CDS) Application Program Interface (API) implemented in
\texttt{\{ecmfwr\}} \citep{Hufkens2019}.

@@

\begin{table}[!ht]
\caption{Description of the earth observation data used }
\label{tab-desEOD}
\small
\centering
\begin{tabular}{p{0.13\textwidth}cp{0.3\textwidth}p{0.095\textwidth}ccc}
\hline
\multirow{1}{*}{\centering Product} & Sub-product & Variable & Spatial Resolution  & Period & Units & Short Name \\ 
\hline
\multirow{4}{*}{ERA5L} & ~ & Precipitation & \multirow{4}{*}{~0.1°} & \multirow{4}{*}{1981-2023} & mm & P \\ 
         &  & Maximum temperature & ~ & & $°C$ & $T_{max}$ \\ 
         &  & Minimum temperature & ~ & & $°C$ & $T_{min}$ \\ 
         &  & Volumetric Soil Water Content at 1m & ~ & & $m3/m3$ & SM \\ 
ERA5L* & & Atmospheric Evaporative Demand & 0.1° & 1981-2023 & mm & AED \\
        \multirow{2}{*}{MODIS} & MOD13A3.061 & Normalized Difference Vegetation Index & \multirow{2}{*}{~1 km} & 2000-2023 & ~ & NDVI \\ 
         & MCD12Q1.061 & Landcover IGBP scheme & & 2001-2022 & ~ & Landcover \\ 
\hline
\end{tabular}
{\raggedright *Derived from ERA5L with Eq. \ref{eq-AED}. \par}
\end{table}

@@

\hypertarget{validation-of-era5l-variables}{%
\subsection{Validation of ERA5L
variables}\label{validation-of-era5l-variables}}

@@

Figure~\ref{fig-zcNDVI_var} and \ref{tab-landcoverTrend} show that the
macrozones with major LULCC for 2001-2022 were ``Centro'', ``Sur'', and
``Austral'' with 36\%, 31\%, and 34\%, respectively. The ``Norte Chico''
shows an increase in barrend land of 111 \(km^2 year^{-1}\) and a
reduction in the class savanna -70 \(km^2 year^{-1}\). In the ``Centro''
and ``Sur'' there are changes in the chilean matorral, with an important
reduction in savanna (-136 to -318 \(km^2 yr^{-1}\)), and an increase in
shrubland and grassland. Showing a change for more dense vegetation
types. It appears to be a change from the surface occupied by cropland
from the ``Centro'' to the ``Sur.'' Also, there is a high increase in
forest (397 \(km^2 yr^{-1}\)) in the ``Sur,'' replacing the savanna
lost.

@@

Regarding vegetation productivity aggregated through the macrozones in
the five landcover macroclasses. In ``Norte Grande'' there is a increase
trend of 0.02 (z-index) per decade, related with types of grassland and
shrubland. There is a negative trend in ``Norte Chico'' with -0.04 and
``Centro'' with -0.02 per decade. In the ``Norte Chico,'' savanna
(-0.05) has the lowest trend, and the rest of the types are around
-0.04. In ``Centro,'' shrubland reaches -0.06, followed by grassland
with -0.05, and croplands and savanna have \textasciitilde-0.01 per
decade. This could be associated either with a reduction in vegetation
surface, a decrease in biomass, or browning \citep{Miranda2023}.
Vegetation reached its lowest values since the year 2019, reaching an
extreme condition in early 2020 and 2022 in the ``Norte Chico'' and
Centro'' (Mega Drought). The ``Sur'' and ``Austral'' show a positive
trend of around 0.016 per decade (Figure~\ref{fig-zcNDVI_var}). The area
most affected by a negative trend in vegetation seems to be associated
with the Chilean matorral \citep{Fuentes2021}, despite the croplands
being as severely impacted by drought as native vegetation does in
``Norte Chico.''.

@@

\hypertarget{impact-for-water-supply-and-demand-and-soil-moisture-in-vegetation-productivity-chequear-cifras-en-el-texto}{%
\subsection{Impact for water supply and demand, and soil moisture in
vegetation productivity (CHEQUEAR CIFRAS EN EL
TEXTO)}\label{impact-for-water-supply-and-demand-and-soil-moisture-in-vegetation-productivity-chequear-cifras-en-el-texto}}

According with what is showed Figure~\ref{fig-corTimeScale},
Figure~\ref{fig-corPerson}, and Table 3. Forest seems to be the most
resistant type to drought, showing that only ``Centro'' is slightly (rsq
= 0.3) impacted by a 12-month soil moisture deficit (SSI-12). Following
by grasslands, which show to be impacted by a SSI-12 with a rsq = 0.4,
and a decrease in water supply (SPI-36 and SPEI-24 with rsq = 0.3) in
the ``Norte Chico'' and to a lesser degree in the ``Norte Grande.'' But,
in the southern macrozones, this type was not affected by water demand,
water supply, or soil moisture. The types that show to be most affected
by variation in climate conditions are shrublands, savannas, and
croplands. For savanna in ``Norte Chico,'' the SSI-12 and SPI-24 reached
an rsq of 0.8 and 0.6, respectively. This value decreases to the south,
but the SSI-12 is still the variable explaining more of the variation in
vegetation productivity (rsq = 0.4 in ``Centro'' and 0.2 in ``Sur''). In
the case of croplands, the SPEI-12, SPI-36, and SSI-12 explain between
40\% and 75\% in ``Centro.'' The most affected type by climatic
variation was shrubland, where in ``Norte Chico'' and ``Centro'' soil
moisture and precipitation explained more than 50\%, with SSI-12 being
the most relevant variable, followed by SPI-36 in ``Norte Chico'' and
SPI-24 in ``Sur.''

\begin{table}[!ht]
\label{tab-corLandcover}
\caption{Summarry per landcover macroclass and macrozone regarding the correlation between zcNDVI with the drought indices EDDI, SPI, SPEI, and SSI for time scales of 1, 3, 6, 12, 24, and 36. The numbers in each cell indicate the time scale that reached the maximum correlation for the landcover and macrozone, and the color indicates the strength of the r-squared obtained with the index and the time scale.}
\includegraphics[]{../output/figs/tabla_r_cor_macro_indice.png}
\end{table}

@@

\hypertarget{validation-of-era5l-variables-1}{%
\subsection{Validation of ERA5L
variables}\label{validation-of-era5l-variables-1}}

@@

1.- Respecto a lo que indica \citep{Vicente-Serrano2018}, de que el
aumento en la tendencia en severidad de la sequía (hidrológica) tiene
que ver más con un aumento de la demanda de agua (LULCC) que a una
tendencia en las condiciones climáticas (SPI-12).

2.- Sobre los tipos de landcover más afectados por los indicadores de
sequía. Asociación con el matorral chileno \citep{Fuentes2021}.
Diferencia entre el Norte Chico y Centro y lo que pasa hacía el sur.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{2}
\tightlist
\item
  Como podrian servir estos resultados para desarrollar un predictor de
  sequía basado.
\end{enumerate}

4.- Qué se podria hacer mejor en futuras investigaciones del tema.

5.- Sequía, aridez, water scarcity. Si puede utilizar el mapa y
proyección climática de Koppen\_Geiger \citep{Beck2023} para dirigir
esta discusión
